---
title: |
  \begin{center}
  \includegraphics[width=0.65\textwidth]{images/logo.png}
  \end{center}
  
  \vspace{1cm}
  \LARGE \textbf{MH4510 Statistical Learning and Data Mining}
  
  \vspace{0.5cm}
  \Large \textbf{Identifying customer segments through Unsupervised Learning}
  
  \vspace{1.5cm}
author: |
  | **Fantastic Four**
  | Choo Yi Ken U2240710B
  | Tong Hao Kit U2240130E
  | Callista Limarta U2240542F
  | Jasmine Teng Shi Min U2240121J
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2: 
    toc: true          
    toc_depth: 3  
  bookdown::html_document2: default
bibliography: references.bib
link-citations: true
csl: advances-in-computational-mathematics.csl
header-includes: 
  - \usepackage{mathpazo}
  - \DeclareMathOperator{\softmax}{softmax}
  - \newcommand{\RR}{\mathbb{R}}
  - \usepackage{float}
  - \usepackage{longtable,booktabs,array}
  - \setlength\LTleft{0pt}
  - \setlength\LTright{0pt}
  - \setlength\LTpre{0pt}
  - \setlength\LTpost{0pt}
  - \usepackage[T1]{fontenc}
  - \usepackage{lmodern}
  - \renewcommand{\rmdefault}{lmr}
  - \renewcommand{\ttdefault}{lmtt}
  - \usepackage[section]{placeins}
  - \usepackage{tocloft}
  - \setlength{\cftbeforesecskip}{2pt}
  - \setlength{\cftbeforesubsecskip}{1pt}
  - \setlength{\cftbeforesubsubsecskip}{0pt}

abstract: |
   \noindent Supply chain networks generate large volumes of transactional data, yet extracting actionable intelligence remains challenging without predefined business rules [@TIWARI2018319]. This contributes to operational inefficiencies, inconsistent delivery performance and poor allocation of logistics resources. To address this real-world challenge, this project applies unsupervised learning to uncover latent behavioural patterns in a global supply-chain dataset containing 180,519 transactions. Three complementary clustering methods – K-means, DBSCAN and Gaussian Mixture Models (GMM) are used to segment customers, detect anomalies and capture overlapping behavioural structures that influence cost and performance. Principal Component Analysis (PCA) is applied to reduce dimensionality while preserving 86.5% of variance, ensuring stable and interpretable clustering outcomes. K-Means identifies six distinct behavioural segments with clear differences in profitability and logistics characteristics (Silhouette = 0.356). DBSCAN revealed rare outliers indicating potential operational errors (Silhouette = 0.541), while GMM captured mixed behavioural tendencies relevant for nuanced customer management (Silhouette = 0.146). A multi-criteria evaluation confirms K-Means as the most balanced and interpretable method. Overall, the results demonstrate how raw operational data can be transformed into strategic insights using unsupervised learning, supporting supply-chain decisions, enabling targeted service differentiation, delivery optimization and risk mitigation.
editor_options: 
  markdown: 
    wrap: 72
---

```{r r global_options, include=FALSE, message=FALSE}
# the following command prevents all the R codes from being included
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      out.width = "60%", out.height = "30%",
                      fig.align='center', fig.width=8, fig.height=5 )

# data manipulation
library(tidyverse) # for manipulation with data

# display
library(knitr)
library(kableExtra)

# data cleaning, feature selection
library(readr)
library(lubridate)
library(janitor)
library(corrplot)
library(car)
library(caret)

# kmeans
library(cluster)
library(factoextra)
library(ggplot2)
library(gridExtra)

# dbscan
library(dbscan)

# gmm
library(mclust)

# cluster evaluation
library(clusterSim)
library(clValid)
library(fpc)
library(plotly)

set.seed(42)
```

\newpage

# Introduction

Modern supply-chain operations generate large volumes of transactional
data, yet organizations lack systematic understanding of customer
heterogeneity in purchasing behaviour, order frequency, profit
contribution and delivery reliability. Without data-driven segmentation,
firms operate with uniform service levels and reactive decision-making,
leading to inefficient resource allocation, inconsistent performance and
limited ability to prioritize high-value customers [@TIWARI2018319].

Unsupervised learning methods — K-Means, DBSCAN and Gaussian Mixture
Models, enable automatic segmentation based on multi-dimensional
behavioural patterns without predefined business rules. This project
applies these techniques to 180,519 global supply-chain transactions to
reveal behavioural groups and anomaly profiles that support strategic
logistics planning, service differentiation and risk mitigation.

## Problem Statement

The machine-learning problem addressed is to automatically segment
customers based on behavioural metrics (order frequency, spending
levels, profit ratios, delivery performance) using unsupervised
clustering. Traditional demographic attributes and manual segmentation
lack discriminatory power and scalability. By uncovering natural
groupings in transactional data, this analysis aims to generate insights
supporting service differentiation, delivery optimization and anomaly
detection.

## Literature Review

Traditional supply-chain segmentation relies on expert judgments or
rule-based thresholds, which overlook complex behavioural relationships
[@Biswas_2016]. Big data analytics enables algorithmic approaches that
reveal interdependencies conventional methods cannot detect
[@TIWARI2018319].

K-Means remains widely applied due to scalability and interpretability
[@JAIN2010651],[@lloyd1982kmeans], but assumes spherical clusters and is
sensitive to initialization. DBSCAN discovers arbitrary-shaped clusters
and identifies noise points representing anomalies or errors
[@ester1996density]. GMM extends flexibility through probabilistic soft
assignments, enabling partial membership across behavioural profiles
[@mclachlan2019finite].

Prior studies demonstrate clustering value in supply-chain contexts
[@GOVINDAN2020101967],[@choi2018bda], but focus on single approach. This
study systematically compares centroid-based, density-based and
model-based methods to evaluate their complementary strengths in
profiling customer behaviour, profitability and delivery performance.

## Objectives

This project aims to derive actionable customer segments that support
supply-chain decision-making. Specifically:

1.  Identify natural customer segments based on behavioural metrics
    without predefined rules.

2.  Compare K-Means, DBSCAN and GMM using weighted evaluation to
    determine the most actionable segmentation.

3.  Detect anomalous profiles indicating fraud, errors or operational
    inefficiencies.

4.  Translate clustering outputs into business insights supporting
    logistics prioritization, service differentiation and risk
    mitigation.

# Data and Methodology

## Data Sources

### Primary Dataset: DataCo SMART Supply Chain

**Dataset Overview:** The DataCo SMART Supply Chain dataset from Kaggle
serves as the primary dataset for this project
(<https://www.kaggle.com/datasets/shashwatwork/dataco-smart-supply-chain-for-big-data-analysis/data>).

The dimensions of the primary dataset is as follows:

```{r echo=FALSE}
df_raw <- read.csv("dataset/DataCoSupplyChainDataset.csv") 
dim(df_raw)

```

**Key Characteristics:** The raw dataset from DataCoSupplyChain is a
transaction-level dataset, where each row represents an entry of
transaction. Table 1 summarizes the dataset and provides a brief
understanding of its size and distribution.

```{r summary-of-data, echo=FALSE}
# Clean order data and shipping date
# Change the order city and country to UTF-8

df_raw <- df_raw %>% mutate(
    # Parse order date
    order_date_clean = parse_date_time(
      order.date..DateOrders.,
      orders = c("mdy HM", "mdy_HM"),
      truncated = 3
    ),
    # Parse shipping date
    shipping_date_clean = parse_date_time(
      shipping.date..DateOrders.,  
      orders = c("mdy HM", "mdy_HM"),
      truncated = 3
    )
  ) %>%
  mutate(
    `Order.City` = iconv(`Order.City`, to = "UTF-8", sub = ""),
    `Order.Country` = iconv(`Order.Country`, to = "UTF-8", sub = ""),
    location = paste(`Order.City`, `Order.Country`, sep = ", ")
  ) %>% dplyr::select(-shipping.date..DateOrders., -order.date..DateOrders.)

summary_stats <- data.frame(
  Metric = c("Total Orders", "Unique Customers", "Countries Served", 
             "Product Categories", "Order Date Range", "Shipping Date Range"),
  Value = c(
    format(nrow(df_raw), big.mark = ","),
    format(length(unique(df_raw$Customer.Id)), big.mark = ","),
    length(unique(df_raw$Order.Country)),
    length(unique(df_raw$Department.Name)),
    paste(
      format(min(df_raw$order_date_clean, na.rm = TRUE), "%B %Y"), 
      "to", 
      format(max(df_raw$order_date_clean, na.rm = TRUE), "%B %Y")
    ),
    paste(
      format(min(df_raw$shipping_date_clean, na.rm = TRUE), "%B %Y"), 
      "to", 
      format(max(df_raw$shipping_date_clean, na.rm = TRUE), "%B %Y")
    )
  )
)
kable(summary_stats, 
      booktabs = TRUE,
      caption = "Summary of DataCoSupplyChain Dataset") %>%  
  kable_styling(latex_options = "HOLD_position")
```

The detailed descriptions of each attribute from the primary dataset is
shown in Appendix.

### Supplementary Dataset: Google Cloud Geocoding API

Since geographic coordinates (latitude/longitude) are necessary for
distance-based analysis, we utilized Google Cloud Geocoding API to
perform geocoding on `Order City`, `Order Country` due to the city being
the finest geographical granularity about the order location.

Each unique city-country pair was concatenated into the standardized
format "`Order City`, `Order Country`" (e.g., "Paris, France", "Tokyo,
Japan") and submitted to the API to retrieve the corresponding latitude
and longitude. The API returns the geographic center (centroid) of each
unique city-country pair (total: 3665 pairs), providing consistent
city-level coordinate precision across all observations.

Then, the geocoded dataset is merged to the primary dataset through a
composite key consisting of `Order City` and `Order Country` through a
one-to-many relationship.

```{r, eval=FALSE}
# geocoding_script.R contains the R code that connects to Google Cloud Backend API to retrieve the latitude and longtitude of City, Country
# source("geocoding_script.R", local = TRUE) # eval is set to FALSE to prevent it from running again since API calls charges
```

```{r include=FALSE}
# we will read the output file instead since we have already run it once
df_geocode <- read.csv("dataset/destinations_geocoded.csv") 
df_combined <- left_join(df_raw, df_geocode, by = c("location" = "location"))
dim(df_combined)
```

## Data Cleaning

We performed the following data cleaning steps on the raw
transactional-level data:

-   Removed columns that contains more than 80% NA values (offer no
    information for modelling).

-   Removed columns that contained only unique values (they have zero
    variance and cannot help to differentiate between customers).

-   Removed records that are duplicated (may inflate sample size and
    skew model results).

-   Removed records not in the correct range such as negative sales or
    negative delivery days (erroneous values that corrupt statistical
    analysis). Specific details are shown in Table 2.

```{r include=FALSE}
# Remove columns 
na_threshold <- 0.80
na_proportions <- colMeans(is.na(df_combined))
cols_to_remove_na <- names(na_proportions[na_proportions > na_threshold])

df_cleaned <- df_combined %>%
  dplyr::select(-all_of(cols_to_remove_na)) %>%
  dplyr::select(where(~n_distinct(., na.rm = TRUE) > 1))
cat("Original number of columns:", ncol(df_combined), "columns. After removing columns:", ncol(df_cleaned), "columns\n\n")

# Remove duplicate rows
n_duplicates <- sum(duplicated(df_cleaned))
cat("Duplicate rows:", format(n_duplicates, big.mark = ","), "\n")
df_cleaned <- df_cleaned %>% distinct()


# Perform type coercion and filter variables
df_cleaned <- df_cleaned %>%
  dplyr::select(-Order.City.y, -Order.Country.y, -location) %>% 
  rename("Order.City"="Order.City.x", "Order.Country"="Order.Country.x", "source_lat"="Latitude", "source_long"="Longitude") %>%
  mutate(
    across(c(Days.for.shipping..real., Days.for.shipment..scheduled., Benefit.per.order, 
             Sales.per.customer, Order.Item.Discount, Order.Item.Discount.Rate, 
             Order.Item.Product.Price, Order.Item.Profit.Ratio, Order.Item.Quantity, Sales, 
             Order.Item.Total, Order.Profit.Per.Order, Product.Price, source_lat, source_long, 
             dest_lat, dest_long, Late_delivery_risk), 
           as.numeric),
    
    across(c(Customer.Id, Order.Id, Order.Customer.Id, Customer.Zipcode, Category.Id, 
             Department.Id, Order.Item.Id, Product.Card.Id, Product.Category.Id, 
             Order.Item.Cardprod.Id), 
           as.character),
    
    across(c(Type, Delivery.Status, Customer.Segment, Department.Name, Market, Shipping.Mode, 
             Category.Name, Order.Status, Order.Region, Customer.City, Customer.Country, 
             Customer.State, Order.City, Order.Country, Order.State), 
           as.factor)
  )

cat("Data types converted successfully\n")

# Number of rows before removing invalid values
n_before <- nrow(df_cleaned) 

df_cleaned <- df_cleaned %>%
  filter(
    # Quantity and prices must be positive
    Order.Item.Quantity > 0,
    Order.Item.Total >= 0,
    Sales >= 0,
    Product.Price >= 0,
    
    # Days can't be negative
    Days.for.shipping..real. >= 0,
    Days.for.shipment..scheduled. >= 0,
    
    # Discount rate should be between 0 and 100
    Order.Item.Discount.Rate >= 0,
    Order.Item.Discount.Rate <= 100,
    
    # Coordinates should be valid ranges
    source_lat >= -90, source_lat <= 90,
    source_long >= -180, source_long <= 180,
    dest_lat >= -90, dest_lat <= 90,
    dest_long >= -180, dest_long <= 180
  )

cat("Invalid records removed:", n_before - nrow(df_cleaned), "\n")

validation_rules <- data.frame(
  Category = c("Quantity & Prices", "", "", "",
               "Shipping Duration", "",
               "Discount Rate", "",
               "Geographic Coordinates", "", "", ""),
  Variable = c("Order.Item.Quantity", "Order.Item.Total", "Sales", "Product.Price",
               "Days.for.shipping..real.", "Days.for.shipment..scheduled.",
               "Order.Item.Discount.Rate", "Order.Item.Discount.Rate",
               "source_lat / dest_lat", "source_lat / dest_lat",
               "source_long / dest_long", "source_long / dest_long"),
  Rule = c("> 0", ">= 0", ">= 0", ">= 0",
           ">= 0", ">= 0",
           ">= 0", "<= 100",
           ">= -90", "<= 90",
           ">= -180", "<= 180"),
  Rationale = c("Quantity must be positive",
                "Total cannot be negative",
                "Sales cannot be negative",
                "Price cannot be negative",
                "Days cannot be negative",
                "Days cannot be negative",
                "Discount rate minimum",
                "Discount rate maximum",
                "Valid latitude minimum",
                "Valid latitude maximum",
                "Valid longitude minimum",
                "Valid longitude maximum")
)
```

```{r include=TRUE}
kable(validation_rules, booktabs = TRUE,
      col.names = c("Category", "Variable", "Rule", "Rationale"),
      caption = "Elimination Rules for Erroneous Datapoints") %>%
  kable_styling(latex_options = "HOLD_position", font_size = 9)
```

## Feature Engineering

Given the dataset's existing complexity with 50 variables, we performed
minimal feature engineering by adding one critical geographic feature:
`delivery_distance_km`, calculated using the Haversine formula to
determine the great-circle distance $d$ between two points on Earth
given their longitude and latitude:

$$d = 2r \arcsin\left(\sqrt{\sin^2\left(\frac{\phi_2 - \phi_1}{2}\right) + \cos(\phi_1)\cos(\phi_2)\sin^2\left(\frac{\lambda_2 - \lambda_1}{2}\right)}\right)$$

```{r include=FALSE}
# definition of haversine distance function
haversine_distance <- function(lon1, lat1, lon2, lat2) {
  # Convert degrees to radians
  lon1_rad <- lon1 * pi / 180
  lat1_rad <- lat1 * pi / 180
  lon2_rad <- lon2 * pi / 180
  lat2_rad <- lat2 * pi / 180
  
  # Differences
  dlon <- lon2_rad - lon1_rad
  dlat <- lat2_rad - lat1_rad
  
  a <- sin(dlat/2)^2 + cos(lat1_rad) * cos(lat2_rad) * sin(dlon/2)^2
  c <- 2 * atan2(sqrt(a), sqrt(1-a))
  
  r <- 6371 # Earth's radius in kilometers
  
  distance <- r * c
  return(distance)
}

df_engineered <- df_cleaned %>%
  mutate(
    delivery_distance_km = haversine_distance(
      source_long, source_lat, dest_long, dest_lat
    )
)

cat("Added 1 new features: delivery_distance_km\n")
cat("Total features:", ncol(df_engineered), "\n")
```

To transition from high-granularity transaction data to low-granularity
customer segments, all relevant numerical transaction variables were
aggregated by `Customer.Id`, resulting in a single behavioural profile
for each customer. This process used sum, mean and count functions to
create features that reflect key aspects of customer interactions with
the supply chain. Numeric variables such as `source_lat`, `source_long`,
`dest_lat` and `dest_long` were excluded as aggregated geographic
coordinates do not provide meaningful insights across multiple
locations. Table 3 summarizes the customer-level feature engineering.

```{r include=FALSE}
numeric_features_cust <- df_engineered %>%
  dplyr::select(where(is.numeric), Customer.Id)

customer_features <- numeric_features_cust %>%
  group_by(Customer.Id) %>% 
  summarise(
    # ===== ORDER BEHAVIOR =====
    n_orders = n(),
  
    # ===== FINANCIAL =====
    total_sales = sum(Sales),
    avg_sales_per_cust = mean(Sales.per.customer),
    avg_order_value = mean(Order.Item.Total),
    
    # ===== PROFIT =====
    avg_profit_per_order = mean(Order.Profit.Per.Order),
    avg_profit_ratio = mean(Order.Item.Profit.Ratio),
    avg_benefit_per_order = mean(Benefit.per.order),
    
    # ===== PRICING =====
    avg_order_item_product_price = mean(Order.Item.Product.Price),
    avg_product_price = mean(Product.Price),
    
    # ===== DISCOUNT BEHAVIOR =====
    tot_discount = sum(Order.Item.Discount),
    avg_discount_rate = mean(Order.Item.Discount.Rate),
    
    # ===== QUANTITY =====
    avg_items_per_order = mean(Order.Item.Quantity),
    
    # ===== DELIVERY PERFORMANCE =====
    avg_days_for_shipping = mean(Days.for.shipping..real.),
    avg_days_scheduled = mean(Days.for.shipment..scheduled.),
    avg_late_delivery = mean(Late_delivery_risk),
    
    # ===== GEOGRAPHIC =====
    avg_delivery_distance = mean(delivery_distance_km),
    
    .groups = "drop"
  ) 

customer_features_no_id = customer_features %>% dplyr::select(-Customer.Id)
```

```{r, include=TRUE}
feature_engineering_summary <- data.frame(
  `Feature Category` = c(
    "Order Behaviour",
    "Financial Contribution",
    "Profitability",
    "Discount/Pricing",
    "Delivery Performance",
    "Logistics Complexity"
  ),
  `Aggregation Function` = c(
    "Count, Mean",
    "Sum, Mean",
    "Mean",
    "Sum, Mean",
    "Mean",
    "Mean"
  ),
  `Example` = c(
    "Number of orders, average items per order",
    "Total sales, average order value",
    "Average profit, profit ratios",
    "Total discount, average discount rate",
    "Shipping days, late-delivery rate",
    "Average delivery distance"
  ),
  `Rationale` = c(
    "Measures engagement frequency and volume.",
    "Measures total spending and typical transaction size.",
    "Measures the average value returned per transaction.",
    "Captures sensitivity to pricing.",
    "Measures the typical fulfillment time and risk.",
    "Measures the logistical challenge of servicing the customer."
  ),
  check.names = FALSE
)

kable(feature_engineering_summary,
      booktabs = TRUE,
      caption = "Summary of Customer-Level Feature Engineering") %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"), 
                font_size = 8.5) %>%
  column_spec(1, bold = TRUE, width = "3cm") %>%
  column_spec(2, width = "2.5cm") %>%
  column_spec(3, width = "4.5cm") %>%
  column_spec(4, width = "5cm")
```

This aggregation reduced the dataset from 180,519 transactions to 20,652
unique customers, with each customer represented by 16 behavioural
metrics spanning the six categories above. These features form the input
for subsequent dimensionality reduction and clustering analysis.

## Exploratory Data Analysis

To obtain an initial overview of the dataset, several engineered
features (e.g., `n_orders`, `total_sales`, `avg_profit_rato`) were
examined and summary statistics including the mean, median, standard
deviation, min value and max value were computed. The results are shown
in Table 4.

```{r include=TRUE}
eda_summary <- customer_features_no_id %>%
  select(n_orders, total_sales, avg_profit_ratio, 
        avg_days_for_shipping, avg_delivery_distance, avg_discount_rate) %>%
  summarise(
    across(everything(), 
           list(
             Mean = ~round(mean(., na.rm = TRUE), 1),
             Median = ~round(median(., na.rm = TRUE), 1),
             SD = ~round(sd(., na.rm = TRUE), 1),
             Min = ~round(min(., na.rm = TRUE), 1),
             Max = ~round(max(., na.rm = TRUE), 1)
           ), .names = "{.col}_{.fn}")
  ) %>%
  pivot_longer(everything(), 
               names_to = c("Feature", ".value"), 
               names_pattern = "(.+)_(.+)")

kable(eda_summary,
      booktabs = TRUE,
      col.names = c("Feature", "Mean", "Median", "SD", "Min", "Max"),
      caption = "Summary Statistics of Selected Features") %>%
  kable_styling(latex_options = "HOLD_position", font_size = 9)
```

We further examined the distributions of the selected customer-level
features to identify possible skewness, outliers and scale differences.
Figure 1 shows the histograms for the 6 selected variables. From the
distribution plots, `n_orders` and `total_sales` exhibit strong
right-skewness, with majority of customers concentrated at lower values.
In contrast, `avg_profit_ratio` shows left-skewness, with most
observations close to zero, indicating that most customers generate
minimal or slightly negative profit margin. Meanwhile, the distributions
of `avg_days_for_shipping`, `avg_delivery_distance` and
`avg_discount_rate` exhibit apporximately normal, suggesting that these
metrics are centered around typical operational values.

```{r include=TRUE, fig.cap="Distributions of Engineered Features"}
customer_features_no_id %>%
  select(n_orders, total_sales, avg_profit_ratio, 
        avg_days_for_shipping, avg_delivery_distance, avg_discount_rate) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "Value") %>%
  ggplot(aes(x = Value)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7, color = "white") +
  facet_wrap(~Feature, scales = "free", ncol = 3) +
  theme_minimal() +
  labs(x = "Value", y = "Frequency") +
  theme(strip.text = element_text(size = 9, face = "bold"))
```

## Feature Selection

To prepare the data for clustering, three feature selection steps were
applied. First, near-zero variance features were removed using
`nearZeroVar` to eliminate variables with little meaningful variation.
Next, a correlation analysis was conducted and features with \|r\|\>0.8
were excluded to reduce redundancies. The resulting correlation matrix
is shown in Figure 2 for reference.

```{r include=FALSE}

# Identify and remove NZV features
nzv <- nearZeroVar(customer_features_no_id, saveMetrics = TRUE)
nzv_features <- rownames(nzv[nzv$nzv == TRUE, ])

cat("Near-zero variance features found:", length(nzv_features), "\n")
if(length(nzv_features) > 0) {
  cat("Removing:", paste(nzv_features, collapse = ", "), "\n")
}

features_after_nzv <- customer_features_no_id %>%
  dplyr::select(-all_of(nzv_features))

cat("Features remaining:", ncol(features_after_nzv), "\n")

```

```{r corr-matrix, include=TRUE, fig.width=20, fig.height=20, fig.cap="Feature Correlation Matrix", fig.pos="H"}
# Calculate correlation matrix
cor_matrix <- cor(features_after_nzv, use = "complete.obs")

corrplot(cor_matrix[1:ncol(features_after_nzv), 1:ncol(features_after_nzv)], 
         method = "color", 
         type = "upper",
         tl.cex = 0.7, 
         tl.col = "black",
         mar = c(0,0,2,0))
```

Table 5 shows the pairwise variables that have correlation greater than
0.8. These variables were removed from the dataset as highly correlated
features can create redundancy and bias distance-based clustering.

```{r, fig.cap="corr-gt-0.8", include=TRUE}
high_cor_pairs <- which(abs(cor_matrix) > 0.80 & abs(cor_matrix) < 1, arr.ind = TRUE)
high_cor_pairs_df <- data.frame(
  Feature1 = rownames(cor_matrix)[high_cor_pairs[,1]],
  Feature2 = colnames(cor_matrix)[high_cor_pairs[,2]],
  Correlation = round(cor_matrix[high_cor_pairs], 3)
) %>%
  filter(Feature1 < Feature2) %>%  # Remove duplicates
  arrange(desc(abs(Correlation)))

 kable(high_cor_pairs_df, booktabs = TRUE, row.names = FALSE,
        col.names = c("Feature 1", "Feature 2", "Correlation"),
        caption = "Pairwise Features with Correlation > 0.80") %>%
    kable_styling(latex_options = "HOLD_position", font_size = 9)
```

```{r include=FALSE}
# Remove features with |correlation| > 0.80
high_cor <- findCorrelation(cor_matrix, cutoff = 0.8, names = TRUE)

cat("\nHighly correlated features (|r| > 0.80):", length(high_cor), "\n")

features_after_cor <- features_after_nzv %>%
  dplyr::select(-all_of(high_cor))

cat("Features remaining:", ncol(features_after_cor), "\n")
```

```{r include=FALSE}
# VIF 

# Calculate VIF
vif_formula <- as.formula(paste(names(features_after_cor)[1], "~ ."))
vif_model <- lm(vif_formula, data = features_after_cor)
vif_values <- vif(vif_model)
  
# Remove VIF > 10
high_vif <- names(vif_values[vif_values > 10])
  
cat("Features with VIF > 10:", length(high_vif), "\n")
  
features_after_vif <- features_after_cor %>%
  dplyr::select(-all_of(high_vif))

cat("Features remaining:", ncol(features_after_vif), "\n")
```

We then further evaluated multicollinearity using Variance Inflation
Factor (VIF) analysis. Following standard practice, features with VIF \>
10 were identified as exhibiting problematic multicollinearity and
removed from the feature set. In our analysis, none of the features
exhibit high VIF, thus no features were removed in this step.

## Principal Component Analysis (PCA)

To address the curse of dimensionality and improve clustering stability,
we apply Principal Component Analysis (PCA) using `prcomp()` function
with scaling enabled. PCA is used to reduce the feature space while
retaining key variance patterns and is particularly beneficial when
features are correlated and exhibit high dimensionality
[@jolliffe2016principal].

```{r, include=FALSE}
set.seed(42)

# Apply PCA
pca_result <- prcomp(features_after_vif, scale. = TRUE)

# Calculate variance explained
variance_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2)
cumulative_variance <- cumsum(variance_explained)
n_pcs <- which(cumulative_variance >= 0.85)[1]
df_pca <- pca_result$x[, 1:n_pcs]

cat("\n=== PCA Summary ===\n")
cat("Original dimensions:", ncol(features_after_vif), "features\n")
cat("Reduced dimensions:", n_pcs, "components\n")
cat("Variance retained:", round(cumulative_variance[n_pcs] * 100, 1), "%\n")
cat("Dimensionality reduction:", round((1 - n_pcs/ncol(features_after_vif)) * 100, 1), "%\n\n")

# Feature loadings for top PCs
cat("Top 3 contributing features per PC:\n")
for(i in 1:min(5, n_pcs)) {
  loadings_pc <- pca_result$rotation[, i]
  top_features <- names(sort(abs(loadings_pc), decreasing = TRUE)[1:3])
  cat("  PC", i, ":", paste(top_features, collapse = ", "), "\n")
}
```

Figure 3 shows the variance explained by each component: variance
concentrates in the first few components, with diminishing returns
beyond PC6. We retain 6 components explaining 86.5% of total variance
using an 85% threshold, reducing dimensionality from
`r ncol(features_after_vif)` features to 6 components while preserving
key variance structure.

```{r scree-plot, include=TRUE, fig.width=10, fig.height=6, fig.cap="PCA Variance Explained", out.width="80%", fig.pos="H"}
# Set up the plot
par(mar = c(5, 5, 4, 5))

# Plot individual variance (left y-axis)
n_show <- length(variance_explained)
plot(1:n_show, variance_explained[1:n_show] * 100, 
     type = "b", pch = 19, col = "steelblue", lwd = 2,
     xlab = "Principal Component", 
     ylab = "Variance Explained (%)",
     main = "Scree Plot with Cumulative Variance",
     cex.lab = 1.2, cex.main = 1.3,
     ylim = c(0, max(variance_explained * 100) * 1.1))
grid()

# Add cumulative variance (right y-axis)
par(new = TRUE)
plot(1:n_show, cumulative_variance[1:n_show] * 100, 
     type = "b", pch = 17, col = "coral", lwd = 2,
     xlab = "", ylab = "", 
     axes = FALSE,
     ylim = c(0, 100))

# Add right y-axis
axis(4, at = seq(0, 100, 20), col.axis = "coral", col = "coral")
mtext("Cumulative Variance Explained (%)", side = 4, line = 3, col = "coral", cex = 1.2)

# Add threshold lines
abline(h = 85, lty = 2, col = "darkgreen", lwd = 1.5)

# Add text
text(n_show * 0.5, 88, "85% threshold", 
     col = "darkgreen", font = 2, cex = 1.0)

# Add legend
legend("right", 
       legend = c("Individual Variance", "Cumulative Variance"),
       col = c("steelblue", "coral"),
       lty = c(1, 1, 2, 2),
       pch = c(19, 17, NA, NA),
       lwd = 2,
       cex = 0.9)

par(mar = c(5, 4, 4, 2))
```

```{r pca-loadings, include=TRUE}
# Create formulas for each PC showing top 3 loadings as percentages
pc_formulas <- data.frame(
  Component = character(),
  Features = character(),
  stringsAsFactors = FALSE
)

for(i in 1:n_pcs) {
  pc_loadings <- pca_result$rotation[, i]
  
  # Get top 3 by absolute value
  top_3_idx <- order(abs(pc_loadings), decreasing = TRUE)[1:3]
  top_3_features <- names(pc_loadings)[top_3_idx]
  top_3_values <- pc_loadings[top_3_idx]
  
  # Convert loadings to percentages (just multiply by 100)
  loading_pct <- round(top_3_values * 100, 1)
  
  # Build string with percentages
  feature_strings <- sapply(1:3, function(j) {
    sign_str <- ifelse(loading_pct[j] >= 0, "+", "")
    paste0(top_3_features[j], "(", sign_str, loading_pct[j], "%)")
  })
  
  features_str <- paste(feature_strings, collapse = ", ")
  
  pc_formulas <- rbind(pc_formulas, 
                       data.frame(Component = paste0("PC", i),
                                  Features = features_str))
}

kable(pc_formulas,
      booktabs = TRUE,
      row.names = FALSE,
      col.names = c("PC", "Top 3 Feature Loadings"),
      caption = "Principal Component Loadings",
      align = c("c", "l")) %>%
  kable_styling(latex_options = "HOLD_position", font_size = 9) %>%
  column_spec(1, bold = TRUE, width = "1.5cm") %>%
  column_spec(2, width = "12.5cm") %>%
  add_footnote("Percentages represent loading coefficients; sign indicates direction")
```

Table 6 reveals the behavioural dimensions captured by each principal
component.

-   **PC1** represents **customer value and geographic reach**: negative
    loadings on financial metrics combined with positive loading on
    delivery distance distinguish low-value, long-distance customers
    (high PC1) from high-value, regional customers (low PC1).

-   **PC2** clearly captures **profitability**, separating
    margin-generating customers from cost-intensive accounts.

-   **PC3** reflects **delivery speed**, with negative loadings
    indicating faster shipping performance (higher PC3 = shorter
    delivery times).

-   **PC4** represents **delivery reliability risk**, with positive
    loading on late deliveries identifying problematic fulfillment
    patterns.

-   **PC5** distinguishes **premium customers** who pay full price for
    high-value orders from discount-dependent buyers.

-   **PC6** captures **strategic high-value discount sensitivity**,
    identifying customers who combine large purchases with negotiated
    pricing.

These six orthogonal dimensions transformed the 16 correlated features
into interpretable behavioural axes for clustering.

```{r include=FALSE}
# Create a copy of the PCA-reduced data for clustering
# This naming convention maintains compatibility with existing code
df_preprocess <- df_pca

cat("\nPreprocessing complete. Ready for clustering.\n")
cat("Dimensions:", dim(df_preprocess), "\n")
```

All subsequent clustering algorithms (K-Means, DBSCAN, GMM) will be
applied to this PCA-reduced feature space rather than the original
scaled features. However, cluster profiling will still be using the
original unscaled features for business interpretability.

# Models

In this section, we introduce the clustering algorithms applied to the
preprocessed dataset.

## K-Means Clustering

K-Means clustering is a centroid-based partitioning algorithm that
groups unlabeled data points into $k$ clusters by minimizing
within-cluster variance. It is widely adopted due to its computational
efficiency [@JAIN2010651].

### Model description

K-Means solves an optimization problem that partitions the dataset into
k disjoint clusters by minimizing the within-cluster sum of squares
(WCSS):

$$\min_{\mathcal{C}} \sum_{i=1}^{k} \sum_{x \in C_i} ||x - c_i||^2$$
where $C_i$ represents the cluster $i$ and
$c_i = \frac{1}{|C_i|} \sum_{x \in C_i} x$ represents the centroid of
cluster $i$

The iterative algorithm that we are specifically using is called Lloyd's
algorithm [@lloyd1982kmeans]. Lloyd's algorithm solves this non-convex
optimization problem through an algorithm called Block Coordinate
Descent (BCD) [@Tang2014Scaling] which minimizes a function by
iteratively updating one block of variables at a time while keeping
others fixed.

**Step 0 (Initialization):** Initialize $t=0$ and cluster centers:
$c_1, c_2,..., c_K$

**Step 1 (Assignment):** Fix centroids $c^{(t-1)}$, optimize
assignments:
$$\delta^{(t)} = \arg\min_{\delta}\sum_{j=1}^{N} \sum_{i=1}^{K} \delta_{ij}(c_i^{t-1} - x_j)^2$$
This assigns each point to its nearest centroid.

**Step 2 (Update):** Fix assignments $\delta^{(t)}$, optimize centroids:
$$c^{(t)} = \arg\min_{c} \sum_{j=1}^{N} \sum_{i=1}^{K} \delta_{ij}^{(t)}(c_i - x_j)^2$$
The centroids are then recalculated as the mean of assigned points:
$$c_i^{(t)} = \frac{\sum_{j=1}^{N} \delta_{ij}^{(t)} x_j}{\sum_{j=1}^{N} \delta_{ij}^{(t)}}$$

Each step monotonically decreases the objective function. However, the
non-convex nature means global optimality is not guaranteed. Despite its
efficiency, K-Means is sensitive to centroid initialization.
Advancements like K-Means++ [@arthur2007kmeansplusplus] improve
initialization by employing a probabilistic approach, while K-Means\|\|
[@bahmani2012scalable] extends this technique to large-scale, parallel
environments.

### Training process and hyperparameter tuning

```{r, include = FALSE}
set.seed(42)

# Elbow Method: Calculate WCSS for k = 1 to 10
wss <- sapply(1:10, function(k) {
  kmeans(df_preprocess, centers = k, nstart = 25, iter.max = 300, 
         algorithm = "Lloyd")$tot.withinss
})

# Silhouette Method: Use ALL customers (no sampling needed!)
sil_scores <- sapply(2:10, function(k) {
  km <- kmeans(df_preprocess, centers = k, nstart = 25, iter.max = 300, 
               algorithm = "Lloyd")
  
  # Calculate silhouette on ALL data
  ss <- silhouette(km$cluster, dist(df_preprocess))
  mean(ss[, 3])
})

elbow_k <- 6
sil_k <- which.max(sil_scores) + 1

cat("Elbow method suggests k =", elbow_k, "\n")
cat("Silhouette method suggests k =", sil_k, "\n\n")

cat("Silhouette scores (all", nrow(df_preprocess), "customers):\n")
for(i in 1:length(sil_scores)) {
  cat("k =", i+1, ": ", round(sil_scores[i], 4), "\n")
}

optimal_k <- sil_k
cat("\n>>> Selected optimal k =", optimal_k, "<<<\n\n")
```

To select the optimal k for our final model, we utilized both elbow
method and silhouette method. The results are shown in Figure 4. Both
elbow method and silhouette method suggest that k=6 is optimal. Hence,
we will use **k=6** to train the final K-Means model with `nstart = 25`
and max iterations = 300.

```{r kmeans-elbow-sil,include=TRUE, fig.width=20, fig.height=5, out.width="100%", fig.cap="Optimal Selection for k", fig.pos="H"}
par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))

# Elbow plot
plot(1:10, wss, type = "b", pch = 19, col = "steelblue", lwd = 2,
     xlab = "Number of Clusters (k)", 
     ylab = "Within-Cluster Sum of Squares",
     main = "Elbow Method",
     cex.lab = 1.1, cex.main = 1.3)
grid()
abline(v = elbow_k, lty = 2, col = "red", lwd = 2)
text(elbow_k + 0.5, max(wss) * 0.9, 
     paste("k =", elbow_k), col = "red", cex = 1.2)

# Silhouette plot
plot(2:10, sil_scores, type = "b", pch = 19, col = "coral", lwd = 2,
     xlab = "Number of Clusters (k)",
     ylab = "Average Silhouette Width",
     main = paste("Silhouette Method"),
     cex.lab = 1.1, cex.main = 1.3)
grid()
abline(v = sil_k, lty = 2, col = "red", lwd = 2)
text(sil_k + 0.5, max(sil_scores) * 0.95, 
     paste("k =", sil_k), col = "red", cex = 1.2)

par(mfrow = c(1, 1))
```

## DBSCAN

Density-Based Spatial Clustering of Applications with Noise (DBSCAN) is
a density-based clustering algorithm that groups unlabeled, closely
packed data points together into several clusters [@ester1996density].
Unlike centroid-based methods (K-Means), DBSCAN can discover clusters of
arbitrary shape and is effective at identifying noise (outliers).

### Model Description

DBSCAN operates by classifying each point in the dataset based on the
density of its local neighborhood, defined by two key hyperparameters:

-   $\epsilon$: The radius defining the $\epsilon$-neighborhood of a
    point. Two points $p,q$ are considered to be neighbours if the
    distance between them, $dist(p,q)\leq\epsilon$

-   **minPts**: The minimum number of data points required to form a
    cluster/core point. It defines the density requirement for forming a
    cluster.

Data points in DBSCAN can be classified into one of the three
categories:

-   **Core Point**: A point $p$ is a core point if its
    $\epsilon$-neighborhood contains at least `minPts` points:
    $$|N_\epsilon(p)|\geq\text{minPts}$$

-   **Border Point**: A point $p$ is a border point if it is not a core
    point, but lies within the $\epsilon$-neighborhood of some core
    point.

-   **Noise Point**: A point that is neither a core point nor a border
    point.

The DBSCAN algorithm identifies clusters through a single-pass
density-based traversal. Unlike iterative optimization methods, DBSCAN
deterministically partitions the dataset by expanding clusters from
high-density seed points, making it a non-iterative, region-growing
approach.

**Step 0 (Initialization):** Initialize cluster counter $k = 0$.

**Step 1 (Core Point Identification):** For each unvisited point $x_i$,
compute the $\epsilon$-neighborhood and identify the core points:
$$N_\epsilon(x_i) = \{x_j \in D \mid \|x_i - x_j\| \leq \epsilon\}$$
**Step 2 (Cluster Expansion):** For each unassigned core point,
increment $k$, assign the core point to cluster $k$, and iteratively add
all density-reachable points.

**Step 3 (Final Assignment):** Assign remaining unassigned points within
$\epsilon$ of a core point as border points, and mark all others as
noise (cluster 0).

The algorithm operates efficiently with a time complexity of
$O(N \log N)$. While DBSCAN is robust to arbitrary shapes, it assumes
clusters have similar density levels. Advanced variants like HDBSCAN
[@campello2013density] and OPTICS [@ankerst1999optics] address these
limitations.

### Training process and hyperparameter tuning

**K-Distance graph for model selection** To select an appropriate
$\epsilon$ value, we employed the k-distance graph method to analyze the
density distribution. Following Sander et al. (1998)
[@sander1998gdbscan], we set $\text{minPts} = 2 \times D = 12$ for
statistical stability in high-dimensional spaces, where $D$ = 6 is the
number of principal components.

```{r include=FALSE}
set.seed(42)

dimensionality <- ncol(df_preprocess)
final_minpts <- 2*dimensionality

kNNdist_results <- list()

knn_dist <- kNNdist(df_preprocess, k = final_minpts)
knn_sorted <- sort(knn_dist)

# Find elbow using second derivative
smooth_dist <- smooth.spline(1:length(knn_sorted), knn_sorted, spar = 0.35)
second_deriv <- diff(diff(smooth_dist$y))
elbow_idx <- which.max(abs(second_deriv)) + 1
elbow_eps <- knn_sorted[elbow_idx]
```

Figure 5 shows the k-distance graph for minPts = 12. The curve reveals
relatively uniform density with a sharp increase beyond point 20,551,
indicating extreme outliers. To identify the elbow computationally, we
apply the **second derivative method**, which detects the point of
maximum curvature—the sharpest bend in the curve where the rate of
change of slope is greatest. This yields $\epsilon$ = 2.6567, marking
the optimal density threshold that separates dense cluster regions from
sparse noise.

```{r k-distance-graph, include=TRUE, fig.width=8, fig.height=5, out.width="70%", fig.cap="K-Distance Graph for Epsilon Selection", fig.pos="H"}
plot(knn_sorted, 
     type = "l", 
     col = "steelblue", 
     lwd = 2,
     main = paste0("K-Distance Graph (minPts = ", final_minpts, ")"),
     xlab = "Points (sorted by k-distance)",
     ylab = paste0(final_minpts, "-Nearest Neighbor Distance"),
     cex.main = 1.2,
     cex.lab = 1.1)
grid(col = "gray90")

# Mark elbow
abline(h = elbow_eps, col = "red", lty = 2, lwd = 2)
abline(v = elbow_idx, col = "red", lty = 3, lwd = 1.5)

text(length(knn_sorted) * 0.6, elbow_eps * 1.15, 
     bquote(epsilon == .(round(elbow_eps, 4))),
     col = "red", cex = 1.1, font = 2)
```

**Epsilon Validation via Silhouette Optimization**

We validated the elbow-suggested $\epsilon$ by testing a range of values
(2.1 to 3.5) and evaluating cluster quality via silhouette scores,
computed on non-noise points only. This two-stage approach ensures
robust parameter selection by balancing geometric density considerations
with empirical cluster quality metrics.

```{r, include=FALSE}
set.seed(42)

# Test epsilon range: 2.1 to 3.5
eps_range <- sort(c(seq(2.1, 3.5, by = 0.2),elbow_eps))

cat("Testing epsilon values from", min(eps_range), "to", max(eps_range), "\n\n")

# Calculate silhouette for each epsilon
sil_scores_eps <- numeric(length(eps_range))
n_clusters_eps <- numeric(length(eps_range))
noise_pct_eps <- numeric(length(eps_range))

for(i in seq_along(eps_range)) {
  eps_val <- eps_range[i]
  
  # Run DBSCAN
  db_result <- dbscan::dbscan(df_preprocess, eps = eps_val, minPts = final_minpts)
  
  n_clusters <- length(unique(db_result$cluster[db_result$cluster != 0]))
  n_noise <- sum(db_result$cluster == 0)
  pct_noise <- (n_noise / nrow(df_preprocess)) * 100
  
  n_clusters_eps[i] <- n_clusters
  noise_pct_eps[i] <- pct_noise
  
  # Calculate silhouette (excluding noise)
  if(n_clusters >= 2 && n_noise < nrow(df_preprocess)) {
    non_noise_idx <- db_result$cluster != 0
    if(sum(non_noise_idx) > n_clusters) {
      sil_scores_eps[i] <- index.S(dist(df_preprocess[non_noise_idx, ]), 
                                    db_result$cluster[non_noise_idx])
    } else {
      sil_scores_eps[i] <- NA
    }
  } else {
    sil_scores_eps[i] <- NA
  }
  
  cat("$\\epsilon$ =", round(eps_val, 2), ": ", n_clusters, "clusters,", 
      round(pct_noise, 1), "% noise, sil =", 
      ifelse(is.na(sil_scores_eps[i]), "N/A", round(sil_scores_eps[i], 3)), "\n")
}

# Find best epsilon by silhouette
valid_idx <- which(!is.na(sil_scores_eps) & n_clusters_eps >= 2 & noise_pct_eps < 20)
if(length(valid_idx) > 0) {
  best_eps_idx <- valid_idx[which.max(sil_scores_eps[valid_idx])]
  best_sil_eps <- eps_range[best_eps_idx]
} else {
  best_sil_eps <- elbow_eps
}

cat("\nK-distance elbow suggests eps =", round(elbow_eps, 3), "\n")
cat("\nSilhouette method suggests eps =", round(best_sil_eps, 3), "\n")

final_eps <- elbow_eps
```

Figure 6 shows the silhouette scores across the tested $\epsilon$ range.
Notably, silhouette optimization achieves its maximum at exactly
$\epsilon$ = 2.657 (marked by the vertical red line), which coincides
precisely with the elbow identified via the second derivative method in
the k-distance graph. This convergence of two independent approaches:
geometric (k-distance curvature) and quality-based (cluster cohesion)
provides strong validation for our parameter selection. We therefore
select $\epsilon$ = 2.657 and **minPts = 12** as optimal parameters for
the final DBSCAN model.

```{r eps-validation, include=TRUE, fig.width=8, fig.height=5, out.width="70%", fig.cap="Epsilon Validation via Silhouette Optimization", fig.pos="H"}
plot(eps_range, sil_scores_eps, 
     type = "b", pch = 19, col = "steelblue", lwd = 2,
     xlab = expression(epsilon), 
     ylab = "Silhouette Score",
     main = paste0("Silhouette-Based Epsilon Selection (minPts = ", final_minpts, ")"),
     cex.lab = 1.1, cex.main = 1.2)
grid()

# Mark optimal epsilon
abline(v = final_eps, lty = 2, col = "red", lwd = 2)
text(final_eps + 0.15, max(sil_scores_eps, na.rm = TRUE) * 0.9, 
     bquote(epsilon[opt] == .(round(final_eps, 4))),
     col = "red", cex = 1.1, font = 2)

# Add annotation for convergence
text(final_eps + 0.15, max(sil_scores_eps, na.rm = TRUE) * 0.8,
     "Matches k-distance elbow",
     col = "red", cex = 0.9)
```

## Gaussian Mixture Model

The Gaussian Mixture Model (GMM) is a probabilistic model-based
clustering algorithm that assumes all data points are generated from a
weighted combination of several Gaussian distributions
[@mclachlan2019finite]. Each Gaussian component represents a cluster,
characterized by its mean ($\mu$) and covariance matrix
($\Sigma$).Unlike hard-assignment methods (K-Means), GMM uses a
soft-assignment method, providing a probability ($\tau_{ik}$) that each
point belongs to each cluster, which is effective for datasets with
overlapping cluster behaviours.

### Model description

GMM models the data distribution as a weighted sum of $K$ Gaussian
components, where the probability density function for a data point
$x_i$ is given by:

$$p(x_i) = \sum_{k=1}^{K} \pi_k \mathcal{N}(x_i \mid \mu_k, \Sigma_k)$$

The goal of GMM is to estimate the parameters
$\theta = \{\pi_k, \mu_k, \Sigma_k\}_{k=1}^{K}$ that maximize the
log-likelihood of the observed data:
$$\mathcal{L}(\theta) = \sum_{i=1}^{N} \log(p(x_i)) $$

A simple formulation of the problem would be deriving the Maximum
Likelihood Estimator (MLE) by maximizing the log-likelihood but the
problem has no closed-form solution due to the summation inside the
logarithm [@mackey2014unsupervised]. Instead, the
Expectation-Maximization (EM) algorithm could provide an iterative
solution by introducing latent variables $\tau_{ik}$ indicating the
probability that point $x_i$ belongs to component $k$.

The EM algorithm iteratively refines the parameters until convergence,
monotonically increasing the log-likelihood.

**Step 0 (Initialization):** Initialize $t=0$ and parameters:
$\theta^{(0)} = (\pi_k^{(0)}, \mu_k^{(0)}, \Sigma_k^{(0)})$ for
$k = 1, ..., K$

**Step 1 (E-step - Expectation):** Compute the posterior probability,
which is a vector of probabilities across all $K$ clusters
[$\tau_{i1}^{(t)}, \tau_{i2}^{(t)}, ..., \tau_{iK}^{(t)}$]. The
probability that $x_i$ belongs to cluster $k$ is given by:

$$\tau_{ik}^{(t)} = \frac{\pi_k^{(t-1)} \mathcal{N}(x_i \mid \mu_k^{(t-1)}, \Sigma_k^{(t-1)})}{\sum_{j=1}^{K} \pi_j^{(t-1)} \mathcal{N}(x_i \mid \mu_j^{(t-1)}, \Sigma_j^{(t-1)})}$$
Note that: $\sum_{k=1}^{K} \tau_{ik}^{(t)} = 1$ for each point $i$.

**Step 2 (M-step - Maximization):** Fix responsibilities
$\tau_{ik}^{(t)}$, update parameters
$\theta^{(t)} = (\pi_k^{(t)}, \mu_k^{(t)}, \Sigma_k^{(t)})$ by
maximizing the expected log-likelihood. The definition of parameters are
given as follows:

-   **Mixing coefficients:**
    $$\pi_k^{(t)} = \frac{1}{N} \sum_{i=1}^{N} \tau_{ik}^{(t)}$$

-   **Means:**
    $$\mu_k^{(t)} = \frac{\sum_{i=1}^{N} \tau_{ik}^{(t)} x_i}{\sum_{i=1}^{N} \tau_{ik}^{(t)}}$$

-   **Covariances:**
    $$\Sigma_k^{(t)} = \frac{\sum_{i=1}^{N} \tau_{ik}^{(t)} (x_i - \mu_k^{(t)})(x_i - \mu_k^{(t)})^T}{\sum_{i=1}^{N} \tau_{ik}^{(t)}}$$

Like K-Means, GMM is sensitive to initialization and requires the number
of clusters ($K$) to be specified. Bayesian GMM (BGMM) [@bishop2006prml]
addresses the latter by utilizing Dirichlet priors to automatically
infer the optimal $K$.

### Training process and hyperparameter tuning

GMM employs two information criteria for model selection:

-   **Bayesian Information Criterion (BIC)**: Balances fit and
    complexity with stronger penalty
    $$\text{BIC} = -2\log(\mathcal{L}) + p\log(N)$$

-   **Integrated Completed Likelihood (ICL)**: Extends BIC by penalizing
    poorly separated clusters
    $$\text{ICL} = \text{BIC} - 2 \times \text{Entropy}$$ where
    $\mathcal{L}$ is the maximized likelihood, $p$ is the number of
    parameters, and $N$ is the sample size.

```{r, include=FALSE}
set.seed(42)

# Test K = 2 to 10
k_range <- 2:10
bic_values <- numeric(length(k_range))
icl_values <- numeric(length(k_range))
model_types <- character(length(k_range))

for(i in seq_along(k_range)) {
  k <- k_range[i]
  gmm_temp <- Mclust(df_preprocess, G = k)
  
  bic_values[i] <- gmm_temp$bic
  icl_temp <- mclustICL(df_preprocess, G = k)
  icl_values[i] <- max(icl_temp, na.rm = TRUE)
  model_types[i] <- gmm_temp$modelName
}

# Find peaks
bic_optimal_idx <- which.max(bic_values)
bic_optimal_k <- k_range[bic_optimal_idx]
icl_optimal_k <- k_range[which.max(icl_values)]

# Select local maximum
optimal_k_gmm <- 6
local_max_idx <- which(k_range == optimal_k_gmm)
optimal_model <- model_types[local_max_idx]

cat("BIC peak: K =", bic_optimal_k, "\n")
cat("ICL peak: K =", icl_optimal_k, "\n")
cat("Selected: K =", optimal_k_gmm, "(", optimal_model, ")\n")
```

**Local Maximum Selection Strategy**

Figure 7 shows BIC and ICL scores across K = 2-10. Both criteria reach
global maxima at K = `r bic_optimal_k`, the upper search limit,
suggesting monotonic increase rather than true convergence. However,
their behavior differs: **BIC exhibits a clear local maximum at K =
`r optimal_k_gmm`** (BIC: `r round(bic_values[local_max_idx], 2)`)
followed by sharp decline at K = 7, while ICL continues increasing
monotonically toward the boundary.

We prioritized BIC's local peak over ICL's monotonic increase. BIC's
decline after K = 6 signals that additional components increase model
complexity without proportionate improvement in fit, suggesting
overfitting. In contrast, ICL's continued increase likely reflects its
entropy term favoring more granular partitions regardless of genuine
structure. The clear inflection point in BIC at K = 6, where the
criterion peaks before declining, provides stronger evidence of natural
clustering structure than ICL's boundary-limited maximum. We therefore
select **K = 6** with **`r optimal_model`** covariance structure, guided
by BIC's local maximum as the more conservative and stable indicator of
true cluster count.

```{r, include=TRUE, fig.width=8, fig.height=5, out.width="70%", fig.cap="GMM Model Selection: BIC and ICL Comparison"}
par(mar = c(4.5, 4.5, 3, 8), xpd = FALSE, fig.pos="H")

# Normalize values to [0, 1] for plotting on same scale
bic_norm <- (bic_values - min(bic_values)) / (max(bic_values) - min(bic_values))
icl_norm <- (icl_values - min(icl_values)) / (max(icl_values) - min(icl_values))

# Create plot
plot(k_range, bic_norm, type = "b", pch = 19, col = "steelblue", lwd = 2.5,
     xlab = "Number of Components (K)",
     ylab = "Normalized Score (Higher is Better)",
     main = "GMM Model Selection: BIC and ICL Comparison",
     ylim = c(0, 1.05),
     cex.lab = 1.1, cex.main = 1.3, cex.axis = 1.0)
grid()

lines(k_range, icl_norm, type = "b", pch = 15, col = "coral", lwd = 2.5)

# Mark optimal points (for reference)
points(bic_optimal_k, bic_norm[bic_optimal_idx], pch = 19, col = "steelblue", cex = 2, lwd = 2)

# Add vertical lines
abline(v = optimal_k_gmm, lty = 2, col = "red", lwd = 2)  # Selected K

# Add text annotation for selected K
text(optimal_k_gmm, 0.5, paste0("Selected\nK = ", optimal_k_gmm), 
     col = "red", font = 2, cex = 1.1, pos = 4)

# Add legend
par(xpd = TRUE)
legend("right", inset = c(-0.25, 0),
       legend = c(
         paste0("BIC"),
         paste0("ICL")
       ),
       col = c("steelblue", "coral"),
       pch = c(19, 15, 19),
       lty = c(1, 1, 1),
       lwd = c(2.5, 2.5, 3),
       cex = 1.0,
       bg = "white",
       box.lwd = 1)
par(xpd = FALSE)
```

# Results

## Overview of Clustering Solutions

Table 7 shows the **overview of clustering solutions** implemented in
this project (K-Means, DBSCAN, GMM). The table summarizes the number of
clusters, approaches and hyperparameter selection method.

```{r, include=TRUE}
# Summary comparison table
clustering_summary <- data.frame(
  Algorithm = c("K-Means", "DBSCAN", "GMM"),
  Clusters = c(6, 2, 6),
  Noise_Points = c(0, 52, 0),
  Noise_Pct = c("0%", paste0(round(0.3, 1), "%"), "0%"),
  Approach = c("Centroid-based", "Density-based", "Model-based"),
  Assignment = c("Hard", "Hard + Noise", "Soft (Probabilistic)")
)

kable(clustering_summary,
      booktabs = TRUE,
      col.names = c("Algorithm", "Clusters", "Noise Points", "Noise %", 
                    "Approach", "Assignment"),
      caption = "Overview of Clustering Solutions") %>%
  kable_styling(latex_options = c("hold_position","striped"),
                full_width = FALSE,      # << stops table from expanding to full width
                position = "center",
                font_size = 9) %>%
  column_spec(1:4, width = "1.5cm") %>%
  # column_spec(2:4, width = "1.5cm") %>%
  column_spec(5:7, width = "2.5cm")
```

## Qualitative Analysis

```{r plot-function, include=FALSE}
# Simple plotting function
plot_pca <- function(clusters, pca_result, title = "") {
  pca_data <- data.frame(
    PC1 = pca_result$x[, 1],
    PC2 = pca_result$x[, 2],
    Cluster = factor(clusters)
  )
  
  centroids <- pca_data %>%
    filter(Cluster != "0") %>%
    group_by(Cluster) %>%
    summarise(PC1 = mean(PC1), PC2 = mean(PC2), .groups = "drop")
  
  # Get colors - gray for cluster 0, Set1 for others
  n_clusters <- length(unique(clusters))
  colors <- c("gray70", RColorBrewer::brewer.pal(max(3, n_clusters-1), "Set1"))
  
  ggplot(pca_data, aes(x = PC1, y = PC2, color = Cluster)) +
    geom_point(alpha = 0.6, size = 1.5) +
    geom_point(data = centroids, aes(x = PC1, y = PC2),
               color = "black", shape = 4, size = 5, stroke = 2) +
    scale_color_manual(values = colors) +
    labs(title = title, x = "PC1", y = "PC2") +
    theme_minimal() +
    theme(legend.position = "bottom",
          plot.title = element_text(hjust = 0.5, face = "bold"))
}
```

```{r train-models, include=FALSE}
set.seed(42)

# K-Means
kmeans_final <- kmeans(df_preprocess, centers = elbow_k, nstart = 25, iter.max = 300)

# DBSCAN
dbscan_final <- dbscan::dbscan(df_preprocess, eps = final_eps, minPts = final_minpts)

# GMM
gmm_final <- Mclust(df_preprocess, G = optimal_k_gmm)
```

```{r all-methods-pca, include=TRUE, fig.width=20, fig.height=6, out.width="100%", fig.cap="Comparison of Clustering Methods in PCA Space", fig.pos="H"}
# Create all three plots
p1 <- plot_pca(kmeans_final$cluster, pca_result, "K-Means")
p2 <- plot_pca(dbscan_final$cluster, pca_result, "DBSCAN")
p3 <- plot_pca(gmm_final$classification, pca_result, "GMM")

# Arrange side by side
grid.arrange(p1, p2, p3, ncol = 3)
```

Figure 8 compares clustering results across all three methods in PCA
space. **K-Means** (left) produces 6 well-separated clusters with points
forming compact clouds around their centroids, demonstrating clear hard
partitioning. **DBSCAN** (center) identifies 2 dense clusters and
explicitly labels 0.3% of customers as noise (gray points), with Cluster
1 containing the majority of customers, indicating concentrated
behavioural patterns. **GMM** (right) reveals 6 Gaussian components with
overlapping boundaries, reflecting probabilistic soft assignment where
customers can have partial membership across segments (model type:
`r gmm_final$modelName`). The comparison highlights each method's
distinct approach: K-Means' geometric partitioning, DBSCAN's
density-based outlier detection and GMM's probabilistic modelling.

## Quantitative Analysis

Clustering performance was assessed using five internal validity indices
that measure the quality of the derived customer segments.

To further determine overall clustering quality, we employ the
Multi-criteria cluster evaluation approach, which aggregates multiple
internal validity indices into a single performance score. This
methodology addresses the challenge of comparing algorithms across
metrics with different scales and interpretations.

The metrics used are: **Silhouette** ($S$), **Davies-Bouldin** ($DB$),
**Calinski-Harabasz** ($CH$), **Dunn** ($D$) and **Variance Explained**
($VE$). Normalized values are scaled to $[0, 1]$; the $DB$ index (lower
is better) is inverted, while other remains higher is better. The
weighted performance score ($\text{Score}_m$) for method $m$ aggregates
the normalized indices:

$$\text{Score}_m = 0.30 \times S_m + 0.25 \times (1-DB_m) + 0.25 \times CH_m + 0.10 \times D_m + 0.10 \times VE_m$$
The weighting scheme was chosen to reflect the primary goal of this
project - to identify actionable and well-separated customer segments.
Therefore, Silhouette score was given the highest priority (30%) as it
measures both cluster cohesion and cluster separation well. Also,
Davies-Bouldin and Calinski-Harabasz scores were also given high
priorities (25% each) as they penalize cluster overlap, which aligns
with the objective of finding distinct customer segments. The remaining
indices (Dunn index and Variance Explained) were given lower priorities
as they represent secondary validation metrics.

```{r eval-function, include=FALSE}
# Function to calculate all evaluation metrics
evaluate_clustering <- function(clusters, data, exclude_noise = FALSE) {
  
  # Handle noise exclusion for DBSCAN
  if(exclude_noise) {
    non_noise_idx <- clusters != 0
    data <- data[non_noise_idx, ]
    clusters <- clusters[non_noise_idx]
  }
  
  # Silhouette Score
  sil_score <- index.S(dist(data), clusters)
  
  # Davies-Bouldin Index
  db_index <- index.DB(data, clusters)$DB
  
  # Calinski-Harabasz Index
  ch_index <- index.G1(data, clusters)
  
  # Dunn Index
  dunn_index <- dunn(distance = NULL, clusters = clusters,
                     Data = data, method = "euclidean")
  
  # Variance Explained (K-Means approximation)
  temp_kmeans <- kmeans(data, centers = max(clusters), nstart = 1)
  var_explained <- (temp_kmeans$betweenss / temp_kmeans$totss) * 100
  
  # Return as named vector
  return(c(
    Silhouette = sil_score,
    Davies_Bouldin = db_index,
    Calinski_Harabasz = ch_index,
    Dunn = dunn_index,
    Var_Explained = var_explained
  ))
}
```

```{r compute-all-metrics, include=FALSE}
# Evaluate all three methods
metrics_kmeans <- evaluate_clustering(kmeans_final$cluster, df_preprocess)
metrics_dbscan <- evaluate_clustering(dbscan_final$cluster, df_preprocess, exclude_noise = TRUE)
metrics_gmm <- evaluate_clustering(gmm_final$classification, df_preprocess)

# DBSCAN noise statistics
n_clusters_dbscan <- length(unique(dbscan_final$cluster[dbscan_final$cluster != 0]))
n_noise_dbscan <- sum(dbscan_final$cluster == 0)
pct_noise_dbscan <- round((n_noise_dbscan / nrow(df_preprocess)) * 100, 1)

# Compile metrics
evaluation_comparison <- data.frame(
  Method = c("K-Means", "DBSCAN", "GMM"),
  K = c(elbow_k, n_clusters_dbscan, optimal_k_gmm),
  Silhouette = c(metrics_kmeans["Silhouette"], 
                 metrics_dbscan["Silhouette"], 
                 metrics_gmm["Silhouette"]),
  Davies_Bouldin = c(metrics_kmeans["Davies_Bouldin"], 
                     metrics_dbscan["Davies_Bouldin"], 
                     metrics_gmm["Davies_Bouldin"]),
  Calinski_Harabasz = c(metrics_kmeans["Calinski_Harabasz"], 
                        metrics_dbscan["Calinski_Harabasz"], 
                        metrics_gmm["Calinski_Harabasz"]),
  Dunn = c(metrics_kmeans["Dunn"], 
           metrics_dbscan["Dunn"], 
           metrics_gmm["Dunn"]),
  Var_Explained = c(metrics_kmeans["Var_Explained"], 
                    metrics_dbscan["Var_Explained"], 
                    metrics_gmm["Var_Explained"])
)

# Weights for multi-criteria evaluation
weights <- c(Silhouette = 0.30, Davies_Bouldin = 0.25, 
             Calinski_Harabasz = 0.25, Dunn = 0.10, Var_Explained = 0.10)

# Normalize and calculate weighted score
eval_norm <- evaluation_comparison %>%
  mutate(
    Sil_norm = (Silhouette - min(Silhouette)) / (max(Silhouette) - min(Silhouette)),
    DB_norm = 1 - ((Davies_Bouldin - min(Davies_Bouldin)) / (max(Davies_Bouldin) - min(Davies_Bouldin))),
    CH_norm = (Calinski_Harabasz - min(Calinski_Harabasz)) / (max(Calinski_Harabasz) - min(Calinski_Harabasz)),
    Dunn_norm = (Dunn - min(Dunn)) / (max(Dunn) - min(Dunn)),
    Var_norm = (Var_Explained - min(Var_Explained)) / (max(Var_Explained) - min(Var_Explained)),
    Score = Sil_norm * weights[1] + DB_norm * weights[2] + 
            CH_norm * weights[3] + Dunn_norm * weights[4] + Var_norm * weights[5]
  )

evaluation_comparison$Score <- round(eval_norm$Score, 4)
```

```{r comparison-table, include=TRUE}
kable(evaluation_comparison,
      booktabs = TRUE,
      digits = c(0, 0, 4, 4, 2, 4, 2, 4),
      col.names = c("Method", "K", "Silhouette", "Davies-Bouldin", 
                    "Calinski-Harabasz", "Dunn", "Var Exp (%)", "Score"),
      caption = "Clustering Performance: Multi-Criteria Evaluation") %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"),
                font_size = 9) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(8, bold = TRUE)
```

Table 8 presents quantitative comparison across all three methods.

K-Means clustering solution with k = 6 is fairly good when evaluated
quantitatively. The Silhouette score (0.3560) indicates good cluster
separation, while the Davies-Bouldin index (1.1456) suggests that the
clusters are moderately separated. The solution explains 56.44% of total
customer variance, demonstrating moderate model fit. It achieved the
highest weighted sum of 0.75 among the three methods, suggesting that
the clustering outcome is comparatively good.

DBSCAN clustering solution achieved a higher raw Silhouette score
(0.5413) than K-Means, indicating that the data points in its two
primary clusters are highly separated and well-defined. Its low
Davies-Bouldin Index (0.9023) further suggests that these clusters are
compact. However, the model explains only 20.04% of the total customer
variance, indicating that this two-cluster solution does not represent
well for the overall structure of the data. The method achieved the
second highest weighted sum of 0.65 among the three methods.

Gaussian Mixture Model (GMM) clustering solution with k = 6 (hard
cluster assignments for evaluation) reports a very low Silhouette score
and high Davies-Bouldin suggesting significant cluster overlap and poor
separation. GMM’s performance reflects its probabilistic soft
boundaries, which are poorly suited for the required crisp customer
segmentation. Also, it achieved the lowest weighted sum of 0.09 among
the three methods.

## Cluster Profiling and Interpretation

```{r}
profile_clusters <- function(cluster_assignments, customer_features, df_engineered,
                             method_name = "Clustering", handle_noise = FALSE) {
  # Add cluster assignments
  customer_clustered <- customer_features %>%
    mutate(Cluster = cluster_assignments)

  # Join to engineered DF 
  df_clustered <- df_engineered %>%
    left_join(customer_clustered %>% dplyr::select(Customer.Id, Cluster),
              by = "Customer.Id")

  # Numerical profiling
  numerical_profiles <- customer_clustered %>%
    group_by(Cluster) %>%
    summarise(
      N_Customers = n(),
      Avg_Orders = round(mean(n_orders, na.rm = TRUE), 1),
      Total_Sales = round(mean(total_sales, na.rm = TRUE), 0),
      Avg_Order_Value = round(mean(avg_order_value, na.rm = TRUE), 2),
      Avg_Profit_Ratio = round(mean(avg_profit_ratio, na.rm = TRUE) * 100, 1),
      Avg_Discount_Rate = round(mean(avg_discount_rate, na.rm = TRUE) * 100, 1),
      Late_Delivery_Risk = round(mean(avg_late_delivery, na.rm = TRUE) * 100, 1),
      .groups = "drop"
    )

  numerical_profiles <- numerical_profiles %>%
    mutate(Pct_Customers = paste0(round(N_Customers / sum(N_Customers) * 100, 1), "%"))

  # Categorical profiling: use df_clustered 
  # Top segment
  top_segment <- df_clustered %>%
    group_by(Cluster, Customer.Segment) %>%
    summarise(n = n(), .groups = "drop_last") %>%
    mutate(pct = n / sum(n) * 100) %>%
    slice_max(pct, n = 1) %>%                
    mutate(Top_Segment = paste0(Customer.Segment, " (", round(pct, 0), "%)")) %>%
    ungroup() %>%
    dplyr::select(Cluster, Top_Segment)

  # Top product
  top_product <- df_clustered %>%
    group_by(Cluster, Category.Name) %>%
    summarise(n = n(), .groups = "drop_last") %>%
    mutate(pct = n / sum(n) * 100) %>%
    slice_max(pct, n = 1) %>%
    mutate(Top_Product = paste0(Category.Name, " (", round(pct, 0), "%)")) %>%
    ungroup() %>%
    dplyr::select(Cluster, Top_Product)

  # Top market
  top_market <- df_clustered %>%
    group_by(Cluster, Market) %>%
    summarise(n = n(), .groups = "drop_last") %>%
    mutate(pct = n / sum(n) * 100) %>%
    slice_max(pct, n = 1) %>%
    mutate(Top_Market = paste0(Market, " (", round(pct, 0), "%)")) %>%
    ungroup() %>%
    dplyr::select(Cluster, Top_Market)

  # Top shipping
  top_shipping <- df_clustered %>%
    group_by(Cluster, Shipping.Mode) %>%
    summarise(n = n(), .groups = "drop_last") %>%
    mutate(pct = n / sum(n) * 100) %>%
    slice_max(pct, n = 1) %>%
    mutate(Top_Shipping = paste0(Shipping.Mode, " (", round(pct, 0), "%)")) %>%
    ungroup() %>%
    dplyr::select(Cluster, Top_Shipping)

  # Combine everything to match your original combined object
  profile <- numerical_profiles %>%
    left_join(top_segment, by = "Cluster") %>%
    left_join(top_product, by = "Cluster") %>%
    left_join(top_market, by = "Cluster") %>%
    left_join(top_shipping, by = "Cluster") %>%
    dplyr::select(Cluster, N_Customers, Pct_Customers, everything())

  profile$Cluster <- as.character(profile$Cluster)

  return(profile)
}
```

```{r generate-profiles, include=FALSE}
# Generate profiles for all methods
profile_kmeans <- profile_clusters(kmeans_final$cluster, customer_features, 
                                   df_engineered, "K-Means")

profile_dbscan <- profile_clusters(dbscan_final$cluster, customer_features, 
                                   df_engineered, "DBSCAN", handle_noise = TRUE)

profile_gmm <- profile_clusters(gmm_final$classification, customer_features, 
                               df_engineered, "GMM")

cat("Profiles generated for all methods\n")
```

```{r kmeans-profile, include=TRUE}
kable(profile_kmeans,
      booktabs = TRUE,
      col.names = c("Cluster", "N", "% Total", "Avg Orders", "Avg Sales ($)", 
                   "Avg Order Value ($)", "Profit %", "Discount %", 
                   "Late Delivery %", "Top Segment", "Top Product", 
                   "Top Market", "Top Shipping"),
      caption = "Customer Segment Profiles: K-Means Clustering") %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"),
                font_size = 8) %>%
  add_header_above(c(" " = 1, "Distribution" = 2, "behavioural Metrics" = 6, 
                    "Demographics" = 4))
```

The six K-Means clusters reflect distinct behavioural patterns. Cluster
1 contains high-value customers who generate negative profits, likely
due to heavy discounts or returns, and require pricing or policy
correction. Cluster 2 is a small but highly profitable premium group
suitable for prioritized service. Cluster 3 represents the core customer
base with frequent purchases and stable profitability. Cluster 4
includes moderate-value customers with high late-delivery rates,
indicating logistics inefficiencies. Cluster 5 consists of low-spending
yet reliable customers with healthy margins. Cluster 6 has moderate
spending but extremely high late-delivery rates, indicating service
issues and potential churn.

Overall, Cluster 1 and Cluster 6 present the most urgent financial and
operational risks, while Clusters 2, 3 and 5 represent stable or
high-value segments that support targeted resource allocation.

```{r dbscan-profile, include=TRUE}
kable(profile_dbscan,
      booktabs = TRUE,
      col.names = c("Cluster", "N", "% Total", "Avg Orders", "Avg Sales ($)", 
                   "Avg Order Value ($)", "Profit %", "Discount %", 
                   "Late Delivery %", "Top Segment", "Top Product", 
                   "Top Market", "Top Shipping"),
      caption = "Customer Segment Profiles: DBSCAN") %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"),
                font_size = 8) %>%
  add_header_above(c(" " = 1, "Distribution" = 2, "behavioural Metrics" = 6, 
                    "Demographics" = 4))
```

Table 10 shows the results of customer profile for DBSCAN. Customers in
Cluster 1 are low-spending consistent buyers. They have high frequency
orders but low order values. On the other hand, Cluster 2 consists of
premium customers. It is a small, very profitable group ideal for
loyalty or premium service programs.

Overall, DBSCAN managed to segment the customers based on order
frequency and value.

```{r gmm-profile, include=TRUE}
kable(profile_gmm,
      booktabs = TRUE,
      col.names = c("Cluster", "N", "% Total", "Avg Orders", "Avg Sales ($)", 
                   "Avg Order Value ($)", "Profit %", "Discount %", 
                   "Late Delivery %", "Top Segment", "Top Product", 
                   "Top Market", "Top Shipping"),
      caption = "Customer Segment Profiles: GMM") %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"),
                font_size = 8) %>%
  add_header_above(c(" " = 1, "Distribution" = 2, "behavioural Metrics" = 6, 
                    "Demographics" = 4))
```

The GMM results reveal six overlapping behavioural segments. Cluster 1
contains low-value customers with extremely high late-delivery rates,
indicating serious service issues. Cluster 2 captures high-value but
unprofitable buyers due to negative margins. Cluster 3 consists of
moderate-value customers who also face high late-delivery risk. Cluster
4 represents high-frequency, profitable customers experiencing poor
delivery reliability, suggesting operational strain. Cluster 5 is the
core and most profitable segment with the highest order frequency.
Cluster 6 includes low-value customers with excellent delivery
performance but limited financial contribution.

Overall, GMM highlights segments facing financial risk (Cluster 2),
service risk (Clusters 1, 3, 4) and strong operational performance
(Cluster 6), with Cluster 5 standing out as the key profit driver.

# Discussion

## Key Findings

The clustering results show that unsupervised learning reveals clear
behavioural segments that directly support supply-chain decision-making.
K-Means (k = 6) produced the most distinct groups, separating customers
by purchasing volume, profitability and delivery reliability, thereby
indicating where to prioritize fulfilment, apply cost-controlled service
levels or address late-delivery risks. DBSCAN identified a small set of
anomalous customers with extreme negative profits, highlighting
potential operational errors important for risk mitigation. GMM captured
overlapping behavioural patterns missed by hard clustering, identifying
customers with mixed characteristics that may benefit from tailored
engagement strategies.

Overall, the multi-criteria evaluation shows K-Means offers the
strongest balance of separation, compactness and interpretability.
Behavioural clustering therefore exposes hidden drivers of cost and
performance, enabling more targeted and efficient supply-chain
decisions.

## Limitations and Future Work

While the clustering analysis produced meaningful behavioural segments,
several limitations remain. First, demographic features such as Top
Segment, Top Product and Top Market showed little variation across
clusters, limiting their usefulness. Future work could engineer richer
attributes such as product diversity or customer lifetime value to
improve discriminatory power. Second, the dataset’s coarse geographic
and product granularity may mask subtle behavioural differences, thus
incorporating finer-grained or derived spatial features would strengthen
segmentation. Third, the algorithms used are sensitive to
hyperparameters. K-Means and GMM require pre-defining the number of
clusters, while DBSCAN depends heavily on $\epsilon$ and minPts. More
robust approaches such as K-Means++, HDBSCAN or BGMM could better
capture complex structures. Finally, the clusters capture associations
rather than causal drivers. Integrating temporal trends or feeding the
segments into downstream predictive models (e.g., late-delivery or churn
prediction) would provide stronger operational value and clearer
evidence of business impact.

# Conclusion

This project addressed the challenge of limited visibility into
heterogeneous customer behaviours that influence cost, service
performance and profitability. By applying unsupervised learning to a
large transactional dataset, we demonstrated how behavioural clustering
can uncover customer segments that are not observable through
traditional reporting.

K-Means produced the most interpretable segmentation, revealing clear
differences in order frequency, spending, profitability and
late-delivery risk, suggesting insights that directly inform fulfilment
prioritization, inventory planning and service-level adjustments. DBSCAN
highlighted anomalous customer patterns associated with potential
operational errors or fraud, while GMM captured overlapping behavioural
tendencies that support more nuanced customer engagement strategies.

Overall, the findings show that unsupervised learning offers a practical
and effective way to transform raw operational data into actionable
segmentation. This data-driven understanding enables more targeted
resource allocation, differentiated service strategies and improved risk
mitigation across supply-chain operations.

\newpage

# References

::: {#refs}
:::

\newpage

# Appendix (Dataset Description)

```{r desc_table, include=TRUE}
col_desc <- read_csv("dataset/DescriptionDataCoSupplyChainDataset.csv")
col_desc <- col_desc %>%
  mutate(FIELDS = gsub("_", "\\\\_", FIELDS)) %>% 
  mutate(FIELDS = paste0("\\textit{", FIELDS, "}")) %>%
  mutate(DESCRIPTION = gsub(":", "", DESCRIPTION)) %>%
  mutate(DESCRIPTION = trimws(DESCRIPTION)) %>%
  mutate(DESCRIPTION = gsub("_", "\\\\_", DESCRIPTION))

kable(col_desc, "latex", 
        escape = FALSE) %>%
    kable_styling(latex_options = c("scale_down", "hold_position"),
                  font_size = 7) %>%
    column_spec(1, width = "3.5cm") %>%  # Fixed width for column 1
    column_spec(2, width = "14cm")     # Fixed width for column 2
```
